---
title: "DANN"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dann}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Package Introduction
DANN is a variation of k nearest neighbors where the shape of the neighborhood takes into account training data's class. The neighborhood is elongated along class boundaries and shrunk in the orthogonal direction to class boundaries. See [Discriminate Adaptive Nearest Neighbor Classification](https://web.stanford.edu/~hastie/Papers/dann_IEEE.pdf)  by Hastie and Tibshirani. This package implements DANN and sub-DANN in section 4.1 of the publication and is based on Christopher Jenness's python [implementation.](https://github.com/christopherjenness/ML-lib)

# Arguments
* k - The number of points in the neighborhood. Identical to k in standard k nearest neighbors.
* neighborhood_size - The number of data points used to estimate a good shape for the neighborhood.
* epsilon - Softening parameter. Usually has the least affect on performance.


# Example: Clustered Data
In this example, a simulated data set is made. There is some overlap between classes.

```{r simData}
library(dann)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(mlbench)

set.seed(1)
train <- mlbench.2dnormals(600, cl = 6, r = sqrt(2), sd = .5) %>%
  tibble::as_tibble()
colnames(train) <- c("X1", "X2", "Y")

ggplot(train, aes(x = X1, y = X2, colour = Y)) + 
  geom_point() + 
  labs(title = "Train Data")


test <- mlbench.2dnormals(600, cl = 6, r = sqrt(2), sd = .5) %>%
  tibble::as_tibble()
colnames(test) <- c("X1", "X2", "Y")
ggplot(test, aes(x = X1, y = X2, colour = Y)) + 
  geom_point() + 
  labs(title = "Test Data")
```

To work with the dann package, data needs to be in matrices instead of dataframes.
```{r Cluster}
xTrain <- train %>%
  select(X1, X2) %>%
  as.matrix()
yTrain <- train %>%
  pull(Y) %>%
  as.numeric() %>%
  as.vector()

xTest <- test %>%
  select(X1, X2) %>%
  as.matrix()
yTest <- test %>%
  pull(Y) %>%
  as.numeric() %>%
  as.vector()
```

To train a model, the matrices and a few parameters are passed into dann. The argument neighborhood_size is the number of data points used to estimate a good shape of the neighborhood. The argument k is the number of data points used in the final classification. Considering there is overlap between all classes and there are only about 100 data points per class, dann performs well for this data set.

```{r model}
dannPreds <- dann(xTrain = xTrain, yTrain = yTrain, xTest = xTest,
                  k = 7, neighborhood_size = 150, epsilon = 1)
round(mean(dannPreds == yTest), 2)
```
