---
title: "Introduction to 'ADtools'"
author: "Chun Fung Kwok"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction-to-ADtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE
)
```

## R package 'ADtools'

Implements the forward-mode automatic differentiation for multivariate functions using the matrix-calculus notation from Magnus and Neudecker (1988). Two key features of the package are: (i) the package incorporates various optimisation strategies to improve performance; this includes applying memoisation to cut down object construction time, using sparse matrix representation to save derivative calculation, and creating specialised matrix operations with Rcpp to reduce computation time; (ii) the package supports differentiating random variable with respect to their parameters, targeting MCMC (and in general simulation-based) applications.

### Installation

```{r, eval = F}
install.packages("ADtools")  # stable version
devtools::install_github("kcf-jackson/ADtools")  # development version
```


---

### Notation

Given a function $f: X \mapsto Y = f(X)$, where $X \in R^{m \times n}, Y \in R^{h \times k}$, the Jacobina matrix of $f$ w.r.t. $X$ is given by
$$\dfrac{\partial f(X)}{\partial X}:=\dfrac{\partial\,\text{vec}\, f(X)}{\partial\, (\text{vec}X)^T} = \dfrac{\partial\,\text{vec}\,Y}{\partial\,(\text{vec}X)^T}\in R^{mn \times hk}.$$

---

### Example 1. Matrix multiplication

#### Function definition
Consider $f(X, y) = X y$ where $X$ is a matrix, and $y$ is a vector. 

```{r}
library(ADtools)
f <- function(X, y) X %*% y
X <- randn(2, 2)
y <- matrix(c(1, 1))
print(list(X = X, y = y, f = f(X, y)))
```

#### Automatic differentiation
Since $X$ has dimension (`r dim(X)`) and $y$ has dimension (`r dim(y)`), the input space has dimension $2 \times 2 + 2 \times 1 = 6$, and the output has dimension $2$, i.e. $f$ maps $R^6$ to $R^2$ and the Jacobian of $f$ should be a $2 \times 6$ matrix.

```{r}
# Full Jacobian matrix
f_AD <- auto_diff(f, at = list(X = X, y = y))
f_AD@dx   # returns a Jacobian matrix
```

`auto_diff` also supports computing a partial Jacobian matrix. For instance, suppose we are only interested in the derivative w.r.t. `y`, then we can run

```{r}
f_AD <- auto_diff(f, at = list(X = X, y = y), wrt = "y")
f_AD@dx   # returns a partial Jacobian matrix
```

#### Finite-differencing

It is good practice to always check the result with finite-differencing. This can be done by calling `finite_diff` which has the same interface as `auto_diff`. 
```{r}
f_FD <- finite_diff(f, at = list(X = X, y = y))
f_FD
```

---

### Example 2. Estimating a linear regression model

#### Simulate data from $\quad y_i = X_i \beta + \epsilon_i, \quad \epsilon_i \sim N(0, 1)$
```{r}
set.seed(123)
n <- 1000
p <- 3
X <- randn(n, p)
beta <- randn(p, 1)
y <- X %*% beta + rnorm(n)
```

#### Inference with gradient descent
```{r}
gradient_descent <- function(f, vary, fix, learning_rate = 0.01, tol = 1e-6, show = F) {
  repeat {
    df <- auto_diff(f, at = append(vary, fix), wrt = names(vary))
    if (show) print(df@x)
    delta <- learning_rate * as.numeric(df@dx)
    vary <- relist(unlist(vary) - delta, vary)
    if (max(abs(delta)) < tol) break
  }
  vary
}
```

```{r}
lm_loss <- function(y, X, beta) sum((y - X %*% beta)^2)

# Estimate
gradient_descent(
  f = lm_loss, vary = list(beta = rnorm(p, 1)), fix = list(y = y, X = X),  learning_rate = 1e-4
) 
# Truth
t(beta)
```


<!-- ### Example  2b. Fitting a 2-layer Neural Network  -->
<!-- #### Simulate data  -->
<!-- ```{r} -->
<!-- logit <- function(x) exp(x) / (1 + exp(x)) -->

<!-- X <- randn(1000, 10) -->
<!-- W1 <- randn(10, 50) -->
<!-- W2 <- randn(50, 1) -->
<!-- f1 <- f2 <- logit -->

<!-- y <- f2(f1(X %*% W1) %*% W2) -->
<!-- ``` -->


<!-- #### Inference with gradient descent -->
<!-- ```{r} -->
<!-- loss_fun <- function(y, X, W1, W2, f1, f2) { -->
<!--   Z <- f1(X %*% W1) -->
<!--   yhat <- f2(Z %*% W2) -->
<!--   sum(y - yhat)^2 -->
<!-- } -->

<!-- gradient_descent( -->
<!--   loss_fun, -->
<!--   vary = list(W1 = W1, W2 = W2), -->
<!--   fix = list(y = y, X = X, f1 = logit, f2 = logit), -->
<!--   learning_rate = 1e-4,  -->
<!--   show = T -->
<!-- ) -->
<!-- ``` -->

---

### Example  3. Sensitivity analysis of MCMC algorithms

#### Simulate data from $\quad y_i = X_i \beta + \epsilon_i, \quad \epsilon_i \sim N(0, 1)$
```{r}
set.seed(123)
n <- 30  # small data
p <- 10
X <- randn(n, p)
beta <- randn(p, 1)
y <- X %*% beta + rnorm(n)
```


#### Estimating a Bayesian linear regression model

$$y \sim N(X\beta, \sigma^2), \quad \beta \sim N(\mathbf{b_0}, \mathbf{B_0}), \quad \sigma^2 \sim IG\left(\dfrac{\alpha_0}{2}, \dfrac{\delta_0}{2}\right)$$

#### Inference using Gibbs sampler
```{r, eval = F}
gibbs_gaussian <- function(X, y, b_0, B_0, alpha_0, delta_0, num_steps = 1e4, burn_ins = ceiling(num_steps / 10)) {
  # Initialisation
  init_sigma <- 1 / sqrt(rgamma0(1, alpha_0 / 2, scale = 2 / delta_0))
  
  n <- length(y)
  alpha_1 <- alpha_0 + n
  sigma_g <- init_sigma
  inv_B_0 <- solve(B_0)
  inv_B_0_times_b_0 <- inv_B_0 %*% b_0
  XTX <- crossprod(X)
  XTy <- crossprod(X, y)
  beta_res <- vector("list", num_steps)
  sigma_res <- vector("list", num_steps)

  pb <- txtProgressBar(1, num_steps, style = 3)
  for (i in 1:num_steps) {
    # Update beta
    B_g <- solve(sigma_g^(-2) * XTX + inv_B_0)
    b_g <- B_g %*% (sigma_g^(-2) * XTy + inv_B_0_times_b_0)
    beta_g <- t(rmvnorm0(1, b_g, B_g))

    # Update sigma
    delta_g <- delta_0 + sum((y - X %*% beta_g)^2)
    sigma_g <- 1 / sqrt(rgamma0(1, alpha_1 / 2, scale = 2 / delta_g))

    # Keep track
    beta_res[[i]] <- beta_g
    sigma_res[[i]] <- sigma_g
    setTxtProgressBar(pb, i)
  }

  # Compute and return the posterior mean
  sample_ids <- (burn_ins + 1):num_steps
  beta_pmean <- Reduce(`+`, beta_res[sample_ids]) / length(sample_ids)
  sigma_pmean <- Reduce(`+`, sigma_res[sample_ids]) / length(sample_ids)
  list(sigma = sigma_pmean, beta = beta_pmean)
}
```

#### Automatic differentiation
```{r, eval = F}
gibbs_deriv <- auto_diff(
  gibbs_gaussian,
  at = list(
    b_0 = numeric(p), B_0 = diag(p), alpha_0 = 4, delta_0 = 4,
    X = X, y = y, num_steps = 50, burn_ins = 5 # Numbers are reduced for CRAN
  ),
  wrt = c("b_0", "B_0", "alpha_0", "delta_0")
)
```

